Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-27 paper-new/discussion.tex
fix trailing sentence
work on discussion

\subsection{Related Work}
% Discuss Joey Eremondi's thesis on gradual dependent types
Eremondi \cite{Eremondi_2023} uses guarded type theory to
define a syntactic model for a gradually-typed source
language with dependent types. By working in guarded type theory, they are
able to give an exact semantics to non-terminating computations in a language
which is logically consistent, allowing for metatheoretic results about the
source language to be proven.
%
Similarly to our approach, they define a guarded lift monad to model potentially-
nonterminating computations and use guarded recursion to model the dynamic type.
However, they do not give a denotational semantics to term precision and it is unclear
how to prove the gradual guarantee in this setting.
The work includes a formalization of the syntactic model in Guarded Cubical Agda.

@@

\subsection{Mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
We will use Guarded Cubical Agda to prove graduality in the syntax of 
GTLC, which involves the construction of the abstract model described in 
\ref{sec:concrete-model} and the extensional model with external dynamic 
type. We also plan to formalize the adequacy result in \ref{sec:appendix-adequacy}.

% step-1
Currently, we have defined the step-1 intensional model in Cubical 
Agda. For value types, we constructed predomain which holds ordering and 
bisimilarity, and morphisms between predmains. For computation types, we 
constructed error domain which contains its underlying predomain, 
elements of $\mho$ and $\theta$ and morphisms between error domains. We 
also defined two functors $U$ and $F$ which support conversions between 
predomains and error domains.

@@

% adequacy
Besides the abstract model and its concrete construction on dynamic type, 
we will also formalize the adequacy result in \ref{sec:appendix-adequacy}, 
which involves clock quantification of the lift monad, the weak bisim 
relation, and the lock-step error ordering. In order to prove adequacy, 
we will first prove that the global lift of X is isomorphic to Delay(1 + X)
whether X is clock-irrelevant or not. Then, we aim to prove the equivalence 
between the global lock-step error ordering and the error ordering observed 
in Delay(1 + X) and equivalence between the global weak bisimilarity 
relation and the weak bisimilarity relation on Delay(1 + X). We have 
finished some prerequisite proofs on clock quantification and postulated 
some theorems on clock globalization.

\subsection{Comparison to Explicit Step-Indexing}

\subsection{Synthetic Ordering}

A key to managing the complexity of our concrete construction is in
using a \emph{synthetic} approach to step-indexing rather than working
analytically with presheaves. This has helped immensely in our ongoing
mechanization in cubical Agda as it sidesteps the need to formalize
these constructions internally. 
%
However, there are other aspects of the model, the bisimilarity and
the monotonicity, which are treated analytically and are .
%
It may be possible to utilize further synthetic techniques to reduce
this burden as well, and have all type intrinsically carry a notion of
bisimilarity and ordering relation, and all constructions to
automatically preserve them.
%
A synthetic approach to ordering is common in (non-guarded) synthetic
domain theory and has also been used for synthetic reasoning for cost
models \cite{synthetic-domain-theory,decalf}.

\subsection{Future Work}

% Cite GrEff paper

>>>>

\subsection{Related Work}
\max{review what I wrote, and discuss Jeremy's work}
% Discuss Joey Eremondi's thesis on gradual dependent types
Eremondi \cite{Eremondi_2023} uses guarded type theory to
define a syntactic model for a gradually-typed source
language with dependent types. By working in guarded type theory, they are
able to give an exact semantics to non-terminating computations in a language
which is logically consistent, allowing for metatheoretic results about the
source language to be proven.
%
Similarly to our approach, they define a guarded lift monad to model potentially-
nonterminating computations and use guarded recursion to model the dynamic type.
However, they do not give a denotational semantics to term precision and it is unclear
how to prove the gradual guarantee in this setting.
The work includes a formalization of the syntactic model in Guarded Cubical Agda.

@@

\subsection{Mechanization}
\max{review}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper,
we are mechanizing our results and developing a reusable framework
for proving graduality in Guarded Cubical Agda \cite{veltri-vezzosi2020}.

As of this writing, the work is in progress, but we have
constructed most parts of the concrete model discussed in
Section \ref{sec:concrete-model}. For instance, we have defined
types for predomains, error domains, and their morphisms and relations,
and we used the guarded features to define the guarded lift + error monad
and the dynamic type.

We plan to formalize the construction of perturbations and 
quasi-representable relations, but we have yet to decide
whether to follow the approach we take in this work and define
the abstract notion of intensional model and formalize the constructions in that setting,
and then apply those abstract constructions to the concrete model.
Alternatively, it may be better from a mechanization standpoint
to carry out those abstract constructions explicitly in the concrete model,
i.e., our representation of objects in the concrete model of predomains
would include a field for the perturbations and our notion of relations
would include fields for the push-pull property and quasi-representability.
We leave this investigation to future work.

Lastly, we would like to formalize the adequacy result discussed in
Section \ref{sec:adequacy}. This will involve adding axioms about
clock quantification as well as about the clock-irrelevance of
booleans and natural numbers, since as of this writing these
axioms are not built-in to Guarded Cubical Agda. As an experiment, we have 
formalized some basic results involving clocks as part of our development.

\begin{comment}
% prove graduality in the syntax of 
% GTLC, which involves the construction of the abstract model described in 
% \ref{sec:concrete-model} and the extensional model with external dynamic 
% type. We also plan to formalize the adequacy result in \ref{sec:appendix-adequacy}.

@@

% adequacy
Besides the abstract model and its concrete construction on dynamic type, 
we will also formalize the adequacy result in \ref{sec:appendix-adequacy}, 
which involves clock quantification of the lift monad, the weak bisim 
relation, and the lock-step error ordering. In order to prove adequacy, 
we will first prove that the global lift of X is isomorphic to Delay(1 + X)
whether X is clock-irrelevant or not. Then, we aim to prove the equivalence 
between the global lock-step error ordering and the error ordering observed 
in Delay(1 + X) and equivalence between the global weak bisimilarity 
relation and the weak bisimilarity relation on Delay(1 + X). We have 
finished some prerequisite proofs on clock quantification and postulated 
some theorems on clock globalization.
\end{comment}

\subsection{Comparison to Explicit Step-Indexing}
\max{check/revise/delete if it's not important}
Working internally to guarded type theory reduces the overhead of needing to
carry around the step-indices in the proofs as is required when using explicit
step-indexing. Additionally, the logical relations constructed to prove
graduality suffer from technical complications of requiring two separate
expression relations, one that counts steps on the left and the other on the right.
There is no analogue of this in our approach.
However, the price we pay for this simplicity is that it is unclear
whether using our approach one can obtain an extensional model that models transitivity
of term precision, which is attainable using the traditional step-indexing approach.

\subsection{Synthetic Ordering}
\max{revise (fix citations) or remove if it's no longer relevant}
A key to managing the complexity of our concrete construction is in
using a \emph{synthetic} approach to step-indexing rather than working
analytically with presheaves. This has helped immensely in our ongoing
mechanization in cubical Agda as it sidesteps the need to formalize
these constructions internally. 
%
However, there are other aspects of the model, the bisimilarity and
the monotonicity, which are treated analytically and are similarly
tedious.
%
It may be possible to utilize further synthetic techniques to reduce
this burden as well, and have all type intrinsically carry a notion of
bisimilarity and ordering relation, and all constructions to
automatically preserve them.
%
A synthetic approach to ordering is common in (non-guarded) synthetic
domain theory and has also been used for synthetic reasoning for cost
models \cite{synthetic-domain-theory,decalf}.

\subsection{Future Work}
\max{review and change if needed}
We would like to apply our approach to give a denotational semantics to gradually-typed
languages with algebraic effects, building on prior work on gradual typing for effect handlers
\cite{greff}. In particular, that work proves graduality via a complicated step-indexed logical relation,
and we hope to prove their results by building a denotational model for GrEff.
This would serve as a step towards applying our techniques to prove graduality for languages
with other advanced features.

The extensional model we construct differs from the usual notion of extensional
model considered in prior work on gradual typing in that it lacks horizontal composition of squares.
We would like to clarify the relationship between our notion of model and prior extensional models,
with the aim of determining whether our approach could allow for the construction of such a model.
